% ============================================================
%  CHAPTER 4 -- Probability Distributions Used in ML
%  Presenter 4
% ============================================================
\section{Probability Distributions Used}

% ---- Slide 1: What is a Probability Distribution? ----
\begin{frame}{What is a Probability Distribution?}
A \textbf{probability distribution} describes how the values of a random variable are spread or distributed.

\begin{columns}[T]
\column{0.48\textwidth}
\begin{block}{Discrete Distribution}
\begin{itemize}
    \item Defined by a \textbf{PMF} $p(x)$
    \item Values are countable
    \item Example: number of heads in 5 coin tosses
\end{itemize}
\end{block}
\column{0.48\textwidth}
\begin{block}{Continuous Distribution}
\begin{itemize}
    \item Defined by a \textbf{PDF} $f(x)$
    \item Values form an interval
    \item Example: height of a person
\end{itemize}
\end{block}
\end{columns}
\vspace{6pt}
\centering
\textit{Every ML model either assumes or learns a probability distribution over the data.}
\end{frame}

% ---- Slide 2: Bernoulli Distribution ----
\begin{frame}{Bernoulli Distribution}
Models a \textbf{single trial} with two outcomes: success ($X=1$) or failure ($X=0$).
\[
X \sim \mathrm{Bernoulli}(p), \quad P(X=1) = p, \;\; P(X=0) = 1-p
\]
\textbf{Properties:}
\begin{itemize}
    \item Mean: $\mathbb{E}[X] = p$
    \item Variance: $\mathrm{Var}(X) = p(1-p)$
\end{itemize}
\begin{exampleblock}{Simple Example}
A coin has $P(\text{Head}) = 0.6$. Then $X \sim \mathrm{Bernoulli}(0.6)$.\\
$\mathbb{E}[X] = 0.6$, \quad $\mathrm{Var}(X) = 0.6 \times 0.4 = 0.24$.
\end{exampleblock}
\textbf{ML usage:} Logistic regression output $\sigma(z) \in (0,1)$ models a Bernoulli probability.
\end{frame}

% ---- Slide 3: Binomial Distribution ----
\begin{frame}{Binomial Distribution}
Sum of $n$ \textbf{independent} Bernoulli trials, each with success probability $p$.
\[
P(X = k) = \binom{n}{k}\, p^k\, (1-p)^{n-k}, \quad k = 0, 1, \dots, n
\]
\textbf{Properties:}
\begin{itemize}
    \item Mean: $\mathbb{E}[X] = np$
    \item Variance: $\mathrm{Var}(X) = np(1-p)$
    \item When $n=1$, it reduces to the Bernoulli distribution.
    \item For large $n$, it approximates a Gaussian (Central Limit Theorem).
\end{itemize}
\begin{exampleblock}{Simple Example}
Toss a fair coin 10 times ($n=10,\; p=0.5$).\\
Expected heads $= np = 5$, \quad Variance $= np(1-p) = 2.5$.
\end{exampleblock}
\end{frame}

% ---- Slide 4: Poisson Distribution ----
\begin{frame}{Poisson Distribution}
Models the \textbf{number of events} occurring in a fixed interval of time or space, when events happen at a constant average rate $\lambda$.
\[
P(X = k) = \frac{\lambda^k\, e^{-\lambda}}{k!}, \quad k = 0, 1, 2, \dots
\]
\textbf{Properties:}
\begin{itemize}
    \item Mean: $\mathbb{E}[X] = \lambda$
    \item Variance: $\mathrm{Var}(X) = \lambda$ \quad (mean = variance!)
    \item Approximates Binomial when $n$ is large and $p$ is small.
\end{itemize}
\begin{exampleblock}{Simple Example}
A server receives $\lambda = 4$ requests per minute on average.\\
$P(\text{exactly 2 requests}) = \frac{4^2\, e^{-4}}{2!} = \frac{16 \times 0.0183}{2} \approx 0.1465$
\end{exampleblock}
\textbf{ML usage:} Count data modelling, event-rate prediction.
\end{frame}

% ---- Slide 5: Uniform Distribution ----
\begin{frame}{Uniform Distribution}
All values in a range are \textbf{equally likely}.

\begin{columns}[T]
\column{0.48\textwidth}
\begin{block}{Discrete Uniform}
\[
P(X = x_i) = \frac{1}{n}
\]
Example: rolling a fair die\\
$P(X=k)=\frac{1}{6}$ for $k \in \{1,\dots,6\}$
\end{block}
\column{0.48\textwidth}
\begin{block}{Continuous Uniform}
\[
f(x) = \frac{1}{b-a}, \quad a \leq x \leq b
\]
Mean: $\frac{a+b}{2}$\\[4pt]
Variance: $\frac{(b-a)^2}{12}$
\end{block}
\end{columns}
\vspace{6pt}
\textbf{ML usage:}
\begin{itemize}
    \item \textbf{Random weight initialisation} in neural networks.
    \item \textbf{Non-informative prior} in Bayesian inference (all parameter values equally likely).
\end{itemize}
\end{frame}

% ---- Slide 6: Gaussian (Normal) Distribution ----
\begin{frame}{Gaussian (Normal) Distribution}
The most important distribution in statistics and machine learning.
\begin{block}{Probability Density Function}
\[
f(x) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\!\left(-\frac{(x-\mu)^2}{2\sigma^2}\right)
\]
\end{block}
\textbf{Key Facts:}
\begin{itemize}
    \item Parameterised by mean $\mu$ (centre) and variance $\sigma^2$ (spread).
    \item Symmetric, bell-shaped curve about $\mu$.
    \item \textbf{68--95--99.7 rule:}
    \begin{itemize}
        \item $\approx 68\%$ of data within $\mu \pm 1\sigma$
        \item $\approx 95\%$ of data within $\mu \pm 2\sigma$
        \item $\approx 99.7\%$ of data within $\mu \pm 3\sigma$
    \end{itemize}
    \item Arises naturally due to the Central Limit Theorem.
\end{itemize}
\end{frame}

% ---- Slide 7: Standard Normal Distribution & Z-Score ----
\begin{frame}{Standard Normal Distribution \& Z-Score}
Any Gaussian $X \sim \mathcal{N}(\mu, \sigma^2)$ can be \textbf{standardised}:
\[
Z = \frac{X - \mu}{\sigma} \sim \mathcal{N}(0, 1)
\]
\begin{itemize}
    \item $Z$ is called the \textbf{standard normal} variable (mean = 0, variance = 1).
    \item Use a \textbf{Z-table} (or calculator) to find probabilities.
    \item $P(X \leq x) = \Phi\!\left(\frac{x - \mu}{\sigma}\right)$, where $\Phi$ is the standard normal CDF.
\end{itemize}
\begin{exampleblock}{Quick Check}
If $X \sim \mathcal{N}(100, 25)$ ($\sigma=5$), what is $P(X \leq 110)$?\\[4pt]
$Z = \frac{110 - 100}{5} = 2.0 \;\Rightarrow\; P(X \leq 110) = \Phi(2.0) = 0.9772$ \quad (97.7\%)
\end{exampleblock}
\textbf{ML usage:} Z-score normalisation ($x' = \frac{x-\mu}{\sigma}$) standardises features before training.
\end{frame}

% ---- Slide 8: Exponential Distribution ----
\begin{frame}{Exponential Distribution}
Models the \textbf{time between successive events} in a Poisson process.
\[
f(x) = \lambda\, e^{-\lambda x}, \quad x \geq 0
\]
\textbf{Properties:}
\begin{itemize}
    \item Mean: $\mathbb{E}[X] = \frac{1}{\lambda}$, \quad Variance: $\mathrm{Var}(X) = \frac{1}{\lambda^2}$
    \item \textbf{Memoryless property:} $P(X > s + t \mid X > s) = P(X > t)$
    \begin{itemize}
        \item The probability of waiting another $t$ units is the same regardless of how long you have already waited.
    \end{itemize}
    \item CDF: $F(x) = 1 - e^{-\lambda x}$
\end{itemize}
\begin{exampleblock}{Simple Example}
If a sensor fails on average every 1000 hours ($\lambda = 0.001$),\\
$P(\text{survives beyond 500 hrs}) = e^{-0.001 \times 500} = e^{-0.5} \approx 0.607$
\end{exampleblock}
\end{frame}

% ---- Slide 9: Summary -- All Distributions at a Glance ----
\begin{frame}{Summary: Distributions at a Glance}
\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{l c c c c}
\toprule
\textbf{Distribution} & \textbf{Type} & \textbf{Mean} & \textbf{Variance} & \textbf{Use Case}\\
\midrule
Bernoulli & Discrete & $p$ & $p(1-p)$ & Binary events\\
Binomial & Discrete & $np$ & $np(1-p)$ & Count of successes\\
Poisson & Discrete & $\lambda$ & $\lambda$ & Rare event counts\\
Uniform & Both & $\frac{a+b}{2}$ & $\frac{(b-a)^2}{12}$ & Initialisation, priors\\
Gaussian & Continuous & $\mu$ & $\sigma^2$ & Noise, features\\
Exponential & Continuous & $\frac{1}{\lambda}$ & $\frac{1}{\lambda^2}$ & Wait times\\
\bottomrule
\end{tabular}
\end{table}
\vspace{4pt}
\textbf{Key takeaway:} Choosing the right distribution to model your data is one of the most important decisions in building any ML/statistical model.
\end{frame}

% ---- Slide 10: Solved Example -- Binomial ----
\begin{frame}{Solved Example: Binomial Distribution}
\begin{exampleblock}{Problem}
A model has 80\% accuracy ($p=0.8$). In a batch of $n=5$ predictions,\\
find $P(\text{exactly 4 correct})$ and $P(\text{at least 4 correct})$.
\end{exampleblock}
\textbf{Solution:}
\begin{align*}
P(X=4) &= \binom{5}{4}(0.8)^4(0.2)^1 = 5 \times 0.4096 \times 0.2 = \mathbf{0.4096}\\[6pt]
P(X=5) &= \binom{5}{5}(0.8)^5(0.2)^0 = 1 \times 0.3277 \times 1 = \mathbf{0.3277}\\[6pt]
P(X \geq 4) &= P(X=4) + P(X=5) = 0.4096 + 0.3277 = \mathbf{0.7373}
\end{align*}
$\Rightarrow$ There is a \textbf{73.7\%} chance of getting 4 or more correct in the batch.
\end{frame}

% ---- Slide 11: Solved Example -- Gaussian & Z-Score ----
\begin{frame}{Solved Example: Gaussian \& Z-Score}
\begin{exampleblock}{Problem}
Feature values follow $X \sim \mathcal{N}(\mu=50,\, \sigma^2=25)$, so $\sigma=5$.\\
Find $P(45 \leq X \leq 60)$.
\end{exampleblock}
\textbf{Solution:} Convert to standard normal $Z = \frac{X-\mu}{\sigma}$:
\begin{align*}
Z_1 &= \frac{45-50}{5} = -1.0, \qquad Z_2 = \frac{60-50}{5} = 2.0\\[6pt]
P(45 \leq X \leq 60) &= P(-1.0 \leq Z \leq 2.0)\\
                       &= \Phi(2.0) - \Phi(-1.0)\\
                       &= 0.9772 - 0.1587 = \mathbf{0.8185}
\end{align*}
$\Rightarrow$ About \textbf{81.9\%} of feature values fall between 45 and 60.
\end{frame}

% ---- Slide 12: Practice Questions ----
\begin{frame}{Practice Questions}
\textbf{Theory Questions:}
\begin{enumerate}
    \item What is the difference between a PMF and a PDF? Give one example of each.
    \item State the 68--95--99.7 rule for the Gaussian distribution. Why is it useful?
    \item Explain the memoryless property of the Exponential distribution in your own words.
    \item When does the Binomial distribution approximate a Gaussian? State the condition.
\end{enumerate}

\vspace{6pt}
\textbf{Numerical Questions:}
\begin{enumerate}
    \item A fair coin is tossed 8 times. Find $P(\text{exactly 5 heads})$. \textit{[Ans: 0.2188]}
    \item If $X \sim \mathcal{N}(70, 16)$ ($\sigma=4$), find $P(X > 78)$. \textit{[Ans: 0.0228]}
    \item On average 3 emails arrive per hour ($\lambda=3$). Find $P(\text{no emails in an hour})$. \textit{[Ans: 0.0498]}
    \item A light bulb lasts on average 200 hours ($\lambda = 0.005$). Find the probability it lasts more than 300 hours. \textit{[Ans: 0.2231]}
\end{enumerate}
\end{frame}
